Active Typechecking and Translation: A Safe Language-Internal Extension Mechanism

### 2013-10-10 esop14/att-esop14.tex
Updated ESOP14

\subsection{Example: Natural Numbers}
\begin{figure}[t]
\small
%\begin{lstlisting}
%family Nat of 1 ~ _.\/(int) {
%	z(i:1, a. case len a =_Z 0 then [[Nat<()> ==> 0]] else err);
%	s(i:1, a. get_last_arg a err (\a1. 
%	  dencase a1 of [[t1 ==> v1]] => 
%	    if t1 =_* Nat<()> then [[Nat<()> ==> ^(v1) + 1]] else err
%	  ow err))
%}
%\end{lstlisting}
\begin{flalign}
&\family{Nat}{\kunit}{\\
&\quad\tops{z}{\kunit}{i}{a}{
	\tapp{\tapp{\tvar{no\_args}}{\tvar{a}}}{
		\tlam{\_}{\kunit}{\tden{\titerm{0}}{\ttype{Nat}{\tunit}}}
	}};\\
&\quad\tops{s}{\kunit}{i}{a}{
	\tapp{
		\tapp{\tvar{last\_arg}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{
			\\
			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{
				\tden{\titerm{\iup{\tvar{v1}}+1}}{\ttype{Nat}{\tunit}}
			}
		}}
		}
	};\\
&\quad\tops{rec}{\kunit}{i}{a}{
	\tapp{\tapp{\tvar{next\_arg}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
	&\quad\quad\tapp{\tapp{\tvar{next\_arg}}{\tvar{a}}}{\tlam{t2}{\kTypeBlur}{\tlam{v2}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
	&\quad\quad \tapp{\tapp{\tvar{last\_arg}}{\tvar{a}}}{\tlam{t3}{\kTypeBlur}{\tlam{v3}{\kITerm}{\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{(\\
		&\quad\quad c
	)}}
	}}}}}}}}}}}}
\\
&}{XXX}{i}{\titype{\dint}}
\end{flalign}
\caption{Example: primitive natural numbers and $n$-ary ($n > 1$) products in \atlam.}
\label{example}
\end{figure}
To make our discussion of each of these components of the calculus concrete, we will begin with an example showing how to introduce primitive natural numbers, implemented as integers internally, as a user-defined extension.

@@

\inferrule[att-op]{
	\fvalDf \in \fvalCtx\\
	\tops{op}{\kappai}{i}{a}{\tauop} \in \theta\\
	\tEvalX{\taui}{\taui'}\\\\
	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\ghat_1}~~~~
	\cdots~~~~
	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\ghat_n}\\
	\text{$\begin{array}{c}
		\left[\text{$\begin{array}{r}
			\taui'/\tvar{i}\\
			\tden{\titerm{\ghat_1}}{\tau_1} :: \cdots :: \tden{\titerm{\ghat_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
			\end{array}$}\right]\tauop
		\Downarrow
		\tden{\titerm{\ghat}}{\ttype{Fam'}{\tauidx}}
		\end{array}$}\\
	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\dbar}\\
	\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}\\
	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\ghat}{\dbar}
}{
	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}{\ttype{Fam'}{\tauidx}}{\ghat}
}
\end{mathpar}
\caption{\small Active typechecking and translation}
\end{figure}
\begin{figure}[t]
\small
\vspace{-15pt}
$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
\begin{mathpar}
\small
\inferrule[var kind]{
}{
  \tKind{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\tvar{t}}{\kappa}
}

@@

\inferrule[deabs i-lam]{
	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\
	\eraseX{\dbar}{\sigma}\\
	\eraseX{\ghat}{\gamma}
}{
	\eraseX{\ilam{x}{\dhat}{\ghat}}{\ilam{x}{\sigma}{\gamma}}
}

%\inferrule[deabs ap]{
%	\eraseX{\ghat_1}{\ghat_1}\\
%	\eraseX{\ghat_2}{\ghat_2}
%}{
%	\eraseX{\iapp{\ghat_1}{\ghat_2}}{\iapp{\ghat_1}{\ghat_2}}
%}
%
\inferrule[deabs i-fix]{
	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\
	\eraseX{\dbar}{\sigma}\\
	\eraseX{\ghat}{\gamma}
}{
	\eraseX{\ifix{x}{\dhat}{\ghat}}{\ifix{x}{\sigma}{\gamma}}
}
\\
\text{\color{gray} (omitted forms have trivially recursive rules)}
\\
\inferrule[deabs unquote]{
	\eraseX{\ghat}{\gamma}
}{
	\eraseX{\iup{\titerm{\ghat}}}{\gamma}
}

@@

CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE 
\section{Discussion}\label{discussion}
\bibliographystyle{abbrv}
\bibliography{../research}
\end{document}

>>>>

\subsection{Example: Natural Numbers}
\begin{figure}[t]
\small
%\begin{lstlisting}
%family Nat of 1 ~ _.\/(int) {
%	z(i:1, a. case len a =_Z 0 then [[Nat<()> ==> 0]] else err);
%	s(i:1, a. get_last_arg a err (\a1. 
%	  dencase a1 of [[t1 ==> v1]] => 
%	    if t1 =_* Nat<()> then [[Nat<()> ==> ^(v1) + 1]] else err
%	  ow err))
%}
%\end{lstlisting}
\begin{flalign}
&\family{Nat}{\kunit}{\\
&\quad\tops{z}{\kunit}{i}{a}{
	\tapp{\tapp{\tvar{const}}{\tvar{a}}}{
		\tlam{\_}{\kunit}{\tden{\titerm{0}}{\ttype{Nat}{\tunit}}}
	}};\\
&\quad\tops{s}{\kunit}{i}{a}{
	\tapp{
		\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{
			\\
			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{
				\tden{\titerm{\iup{\tvar{v1}}+1}}{\ttype{Nat}{\tunit}}
			}
		}}
		}
	};\\
&\quad\tops{rec}{\kunit}{i}{a}{
	\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{t1}{\kTypeBlur}{\tlam{v1}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{t2}{\kTypeBlur}{\tlam{v2}{\kITerm}{\tlam{a}{\klist{\kDen}}{\\
	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{t3}{\kTypeBlur}{\tlam{v3}{\kITerm}{\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
		&\quad\quad \tden{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
		&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{v2}}}{\iapp{\iapp{\iup{\tvar{v3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{v1}}}}{\tvar{t2}}
	)}}
	}}}}}}}}}}}}
\\
&}{XXX}{i}{\titype{\dint}};\\
&\pdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
&(\elam{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat}))}}}{
	\elam{two}{\tvar{nat}}{\\
	&\quad\quad \eopapp{\eopapp{plus}{two}}{two}}})
}\\
&\quad \elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
	\eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
	\eop{Nat}{s}{\tunit}{r}
	}}}}}\\
&\quad  \eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}
%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
%}
%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
	%\eop{Nat}{s}{\tunit}{r}
	%}}}}
\end{flalign}
\caption{Example: primitive natural numbers in \atlam. Type-level helper functions $\tvar{const}$, $\tvar{pop}$, $\tvar{pop\_final}$ and $\tvar{check\_type}$ are given in Appendix \ref{helper}.}
\label{example}
\end{figure}
To make our discussion of each of these components of the calculus concrete, we will begin with an example showing how to introduce primitive natural numbers, implemented as integers internally, as a user-defined extension.

@@

\inferrule[att-op]{
	\fvalDf \in \fvalCtx\\
	\tops{op}{\kappai}{i}{a}{\tauop} \in \theta\\
	\tEvalX{\taui}{\taui'}\\\\
	\ecompiles{\etCtx}{\fvalCtx}{e_1}{\tau_1}{\ghat_1}~~~~
	\cdots~~~~
	\ecompiles{\etCtx}{\fvalCtx}{e_n}{\tau_n}{\ghat_n}\\
	\text{$\begin{array}{c}
		\left[\text{$\begin{array}{r}
			\taui'/\tvar{i}\\
			\tden{\titerm{\ghat_1}}{\tau_1} :: \cdots :: \tden{\titerm{\ghat_n}}{\tau_n} :: \tnil{\kDen}/\tvar{a}
			\end{array}$}\right]\tauop
		\Downarrow
		\tden{\titerm{\ghat}}{\ttype{Fam'}{\tauidx}}
		\end{array}$}\\
	\tauisdel{Fam}{\fvalCtx}{Fam'}{\tauidx}{\dbar}\\
	\eCtxTogCtx{Fam}{\fvalCtx}{\etCtx}{\gtCtx}\\
	\checkRC{\gtCtx}{Fam}{\fvalCtx}{\ghat}{\dbar}
}{
	\ecompiles{\etCtx}{\fvalCtx}{\eop{Fam}{op}{\taui}{\splat{e}{1}{n}}}{\ttype{Fam'}{\tauidx}}{\ghat}
}
\end{mathpar}
\caption{\small Active typechecking and translation}
\end{figure}
\begin{figure}[t]
\small
%\vspace{-15pt}
$\fbox{\inferrule{}{\tKindX{\tau}{\kappa}}}$
\begin{mathpar}
\small
\inferrule[var kind]{
}{
  \tKind{\tvarCtxX{t}{\kappa}}{\fvarCtx}{\tvar{t}}{\kappa}
}

@@

\inferrule[deabs i-lam]{
	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\\\
	\eraseX{\dbar}{\sigma}\\\\
	\eraseX{\ghat}{\gamma}
}{
	\eraseX{\ilam{x}{\dhat}{\ghat}}{\ilam{x}{\sigma}{\gamma}}
}

%\inferrule[deabs ap]{
%	\eraseX{\ghat_1}{\ghat_1}\\
%	\eraseX{\ghat_2}{\ghat_2}
%}{
%	\eraseX{\iapp{\ghat_1}{\ghat_2}}{\iapp{\ghat_1}{\ghat_2}}
%}
%
\inferrule[deabs i-fix]{
	\ddbar{\_}{\fvalCtx}{\dhat}{\dbar}\\\\
	\eraseX{\dbar}{\sigma}\\\\
	\eraseX{\ghat}{\gamma}
}{
	\eraseX{\ifix{x}{\dhat}{\ghat}}{\ifix{x}{\sigma}{\gamma}}
}
\\
\text{\color{gray} (omitted forms have trivially recursive rules)}
\\
\inferrule[deabs unquote]{
	\eraseX{\ghat}{\gamma}
}{
	\eraseX{\iup{\titerm{\ghat}}}{\gamma}
}

@@

CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE 
\section{Discussion}\label{discussion}
\bibliographystyle{abbrv}
\bibliography{../research}
\appendix
\section{Appendix}
\subsection{Helper Functions}\label{helper}

\end{document}
